$\pi$ with leftovers: \\ a mechanisation in Agda

### 2021-01-25 paper/paper.tex
Get rid of structural congruence and reduction subsections

We start by using type level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
We then provide an operational semantics for the \picalc{}, prior to any typing (\autoref{semantics}).
This operational semantics is defined as a reduction relation on processes (\autoref{operational-semantics}).
The reduction relation tracks at the type level the channel on which communication occurs.
This information is later used to state the subject reduction theorem --- aka type preservation.
The reduction relation is defined modulo \emph{structural congruence} (\autoref{structural-congruence}) --- a relation defined on processes that acts as a quotient type to remove unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
We then define an interface for resource-aware algebras (\autoref{multiplicities}) and use it to parametrise a type system based on leftover typing (\autoref{leftover-typing}).
Finally, we present the meta theoretical properties of our \picalc{} with leftovers in \autoref{meta-theory}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Operational Semantics}
\label{semantics}
Thanks to our well-scoped grammar in \autoref{syntax}, the semantics of our language can now be defined on the totality of the syntax.
We define structural congruence in \autoref{structural-congruence} and provide a reduction relation in \autoref{operational-semantics}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Structural Congruence}
\label{structural-congruence}

@@

\begin{nidefinition}[\textsc{StructCong}]
We define the base cases of a structural congruence relation $\eqeq$ as follows:
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

\begin{nidefinition}[\textsc{Equals}]
We lift the relation \textsc{StructCong} $\eqeq{}$ and close it under equivalence and congruence in $\eq{}$.
This relation is structurally congruent under a context $\mathcal{C}[\cdot]$ \cite{Sangio01} and is reflexive, symmetric and transitive. 
\end{nidefinition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}
\label{operational-semantics}

\begin{nidefinition}[\textsc{Reduces}]
The operational semantics of the \picalc{} is defined as a reduction relation $\reduce{c}$ indexed by the channel $c$ on which communication occurs.
We keep track of channel $c$ so we can state subject reduction (\autoref{thm:subject-reduction}).

@@

\begin{nidefinition}[\textsc{Algebra}]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  Below we define such an algebra as a record $\Algebra_C$ on a carrier $C$.
  We use $\forall$ for universal quantification.
  The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
  The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.
  \begin{equation*}
  \begin{aligned}
    &\zero                   &:{} &                 &        & C \\
    &\one                    &:{} &                 &        & C \\
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \\
    &                        &    &                  &        & \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
\end{nidefinition}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[\textsc{Algebras}]
  An \emph{indexed set of usage algebras} is a type $\Idx$ of indices that is nonempty ($\type{\exists IDX}$) together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras of the corresponding type.
  \begin{equation*}
    \begin{aligned}
      &\Idx               &:{} &\Set \\
      &\type{\exists IDX} &:{} &\Idx \\
      &\Usage             &:{} &\Idx \to \Set \\
      &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
    \end{aligned}
  \end{equation*}
\end{nidefinition}

@@

Subject congruence states that applying structural congruence (\autoref{structural-congruence}) to a well-typed process preserves its well-typedness.
Subject reduction states that a well-typed process that communicates on a channel $c$ (\autoref{operational-semantics}) preserves its well-typedness and consumes $\lio$ from $c$ if $c$ is a channel in the environment.

>>>>

We start by using type level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
We then provide an operational semantics for the \picalc{}, prior to any typing (\autoref{semantics}).
This operational semantics is defined as a reduction relation on processes.
The reduction relation tracks at the type level the channel on which communication occurs.
This information is later used to state the subject reduction theorem --- aka type preservation.
The reduction relation is defined modulo \emph{structural congruence} --- a relation defined on processes that acts as a quotient type to remove unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
We then define an interface for resource-aware algebras (\autoref{multiplicities}) and use it to parametrise a type system based on leftover typing (\autoref{leftover-typing}).
Finally, we present the meta theoretical properties of our \picalc{} with leftovers in \autoref{meta-theory}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Operational Semantics}
\label{semantics}
Thanks to our well-scoped grammar in \autoref{syntax}, the semantics of our language can now be defined on the totality of the syntax.

@@

\begin{nidefinition}[\textsc{StructCong}]
  \label{def:structcong}
We define the base cases of a structural congruence relation $\eqeq$ as follows:
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

\begin{nidefinition}[\textsc{Equals}] \label{equals}
We lift the relation \textsc{StructCong} $\eqeq{}$ and close it under equivalence and congruence in $\eq{}$.
This relation is structurally congruent under a context $\mathcal{C}[\cdot]$ \cite{Sangio01} and is reflexive, symmetric and transitive. 
\end{nidefinition}

\begin{nidefinition}[\textsc{Reduces}]\label{def:reduces}
The operational semantics of the \picalc{} is defined as a reduction relation $\reduce{c}$ indexed by the channel $c$ on which communication occurs.
We keep track of channel $c$ so we can state subject reduction (\autoref{thm:subject-reduction}).

@@

\begin{nidefinition}[\textsc{Algebra}]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  Below we define such an algebra as a record $\Algebra_C$ on a carrier $C$.
  We use $\forall$ for universal quantification.
  The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
  The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.

  \begin{equation*}
  \begin{aligned}
    &\zero                   &:{} &                 &        & C \\
    &\one                    &:{} &                 &        & C \\
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \\
    &                        &    &                  &        & \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
\end{nidefinition}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[\textsc{Algebras}]
  An \emph{indexed set of usage algebras} is a type $\Idx$ of indices that is nonempty ($\type{\exists IDX}$) together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras of the corresponding type.

  \begin{equation*}
    \begin{aligned}
      &\Idx               &:{} &\Set \\
      &\type{\exists IDX} &:{} &\Idx \\
      &\Usage             &:{} &\Idx \to \Set \\
      &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
    \end{aligned}
  \end{equation*}
\end{nidefinition}

@@

Subject congruence states that applying structural congruence (\autoref{def:equals}) to a well-typed process preserves its well-typedness.
Subject reduction states that a well-typed process that communicates on a channel $c$ (\autoref{def:reduces}) preserves its well-typedness and consumes $\lio$ from $c$ if $c$ is a channel in the environment.
